Left-linear rules in adhesive categories

### 2024-02-06 icalp/fusione.tex
minima

One of the key pay-off of concurrency theory is the idea that the behaviour of 
a computational device can be modelled by abstracting its sequences of steps
%performed by the device 
via a suitable equivalence. Intuitively, this equivalence 
captures when steps are causally unrelated and thus could be executed in any 
order, and possibly concurrently. 
%
Two of the seminal contribution to this line of research have been
Mazurkiewicz's traces~\cite{xxx} and Winskel's event structures~\cite{xxx}
%
Concerning formalisms based on the rewriting paradigm, i.e. where the
behaviour is dictated by a set of rules which have to be instantiated with
the system at hand, concurrency often boils down to having a notion of 
independence between two consecutive steps.
%thus viewing a sequence of steps as a concurrent computation. 
As noteworthy examples, we have 
interchange law~\cite{Mes92} in term rewriting and permutation 
equivalence~\cite{JJL80} in $\lambda$-calculi.

Among rewriting-based formalisms, those manipulating graph-like structures
proved useful in many settings for representing the dynamics of systems: the 
graph serves to represent the entities and their relations within states, while 
rewriting steps, modifying the graph, model computation steps that results 
in changes to the system state. The DPO (double-pushout) approach~\cite{xxx} is
nowadays considered the standard one for these structures, mostly thanks to its 
flexibility: the rules allow to specify that, 
in a rewriting step, some entities in the current state are removed, some others are 
needed for the rewriting step to occur but remain unaltered and some new entities 
can be generated. Concerning concurrency, the chosen notion is shift equivalence,
and independence is formally captured by the so-called Church-Rosser 
theorem of DPO rewriting~\cite{CMREHL:AAGT}.

>>>>

One of the key pay-off of concurrency theory is the idea that the behaviour of 
a computational device can be modelled by abstracting its sequences of steps
%performed by the device 
via a suitable equivalence. Intuitively, this equivalence 
captures when steps are causally unrelated and thus could be executed in any 
order, and possibly concurrently. 
%
Two seminal contributions to this line of research have been
Mazurkiewicz's traces~\cite{Mazurkiewicz86} and Winskel's event structures~\cite{NPW:PNES}
%
Concerning formalisms based on the rewriting paradigm, i.e. where the
behaviour is dictated by a set of rules which have to be instantiated with
the system at hand, concurrency often boils down to having a notion of 
independence between two consecutive steps.
%thus viewing a sequence of steps as a concurrent computation. 
As noteworthy examples, we have 
interchange law~\cite{Mes92} in term rewriting and permutation 
equivalence~\cite{JJL80} in $\lambda$-calculi.

Among rewriting-based formalisms, those manipulating graph-like structures
proved useful in many settings for representing the dynamics of systems: the 
graph serves to represent the entities and their relations within states, while 
rewriting steps, modifying the graph, model computation steps that results 
in changes to the system state. The DPO (double-pushout) approach~\cite{xxx} is
nowadays considered the standard one for these structures, thanks to its localality 
(rule application has a local effect on a state, differently e.g. from the SPO and SqPO 
approaches~\cite{xxx,xxx}) and most importantly to its flexibility: the rules allow to 
specify that,  in a rewriting step, some entities in the current state are removed, 
some others are needed for the rewriting step to occur but remain unaltered and 
some new entities can be generated. Concerning concurrency, the chosen notion 
is shift equivalence, and independence is formally captured by the so-called 
Church-Rosser theorem of DPO rewriting~\cite{CMREHL:AAGT}.
